# Git Fundamentals {.tabset}

*March 13, 2017*

## Article

### Overview

Of all the tools I use as a data scientist, the one that I cherish the most is **[Git](https://git-scm.com/)**. As a free and open source *distributed* version control system, Git plays an integral role in my work by seamlessly fostering many of the most important considerations of data science workflows including **collaboration**, **experimentation**, **reproducibility** and of course, **source code management**.

At their core, version control systems (VCS) all serve one broad and common purpose: *tracking changes to files*. What distinguishes one system from another, however, is *how* that purpose is implemented and what additional features are present. To get a better understanding of the history and evolution of version control systems, I recommend reading the introduction at [Ry's Git Tutorial](http://rypress.com/tutorials/git/index). You might also want to bookmark his tutorial and work your way through all of the sections as he does a terrific job demonstrating Git's feature set in much greater detail than what I'm covering in this post.

Before we move on, let's clarify something that often comes up when people first start learning about Git. Many of you may have heard of [GitHub](https://github.com/) before, and if you've read through [my resources page](/resources) you may have caught notice of [GitLab](https://about.gitlab.com/) as well. These are each examples of *web-based repository hosting services*. Git itself is just a lightweight command line tool. Services like GitHub provide development platforms that center around the use of Git but add a rich suite of additional features which are useful in software development. The focus of this post is on learning the fundamentals of the Git command line tool.

### Prerequisites

A basic understanding of Linux (Unix-like) commands is assumed. Specifically, this post makes extensive use of the following:

* `mkdir`: create new directories
* `cd`: change the current working directory
* `echo`: display a line of text
* `cat`: concatenate files and print on the standard output
* `ls`: list directory contents
* `rm`: remove files or directories

For brevity, I am using `echo` with redirection (`>`) to write files. In reality, you'd be using a visual editor such as `vi` to write and edit files.

### Create a Local Git Repository

Before we can do anything with Git, we must initialize a directory as a Git repository. Let's do so in a brand new directory that we'll create underneath our home directory (`~`) called `fundamentals`:

```{r, echo=FALSE}
knitr::opts_chunk$set(comment=NA)
knitr::opts_knit$set(root.dir='/home/dnegrey')
```

```{bash}
cd ~
mkdir fundamentals
cd fundamentals
git init
```

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir='/home/dnegrey/fundamentals')
```

### Add a README File

The first thing you'll want to do with any new Git repository is add a **README.md** file to the project root. As a plain text file, it will be the easiest place to save and read notes about your project. The ".md" extension indicates that it is a [markdown](https://daringfireball.net/projects/markdown/) file. Markdown is a text-to-HTML conversion tool that allows you to create easy-to-read and easy-to-write plain text files which get converted to HTML. Hosting services like GitHub and GitLab will automatically render your README.md file to HTML at the main repository site (for example: [https://github.com/dnegrey/blogr](https://github.com/dnegrey/blogr)). Here is a brief primer on some more commonly used markdown syntax:

---

<div>
<div style="float: left; width: 48%; margin-left: 0; margin-right: 2%;">

##### **Headers**

```{bash, eval=FALSE, highlight=FALSE}
# is an h1 header
## is an h2 header
### is an h3 header (and so on)
```

##### **Regular Writing**

```{bash, eval=FALSE, highlight=FALSE}
Regular writing becomes a <p> tag
```

##### **Inline Code**

```{bash, eval=FALSE, highlight=FALSE}
Enclose inline code in `single ticks`
```

##### **Unordered Lists**

```{bash, eval=FALSE, highlight=FALSE}
* item one in unordered list
* item two in unordered list
* item three in unordered list
```

</div>
<div style="float: right; width: 48%; margin-left: 2%; margin-right: 0;">

##### **Italics and Bold**

```{bash, eval=FALSE, highlight=FALSE}
*italics*
**bold**
***bold-and-italics***
```

##### **Hyperlinks**

```{bash, eval=FALSE, highlight=FALSE}
[hyperlink-alt-text](hyperlink-href)
```

##### **Code Chunks**

```{bash, eval=FALSE, highlight=FALSE}
    Indent four spaces for a code block
```

##### **Ordered Lists**

```{bash, eval=FALSE, highlight=FALSE}
1 item one in ordered list
2 item two in ordered list
3 item three in ordered list
```

</div>
<br style="clear:both;"/>
</div>

---

Now let's actually create our README.md file. A popular convention (and one that I use) is to have the first line of your README.md file be an `<h1>` header with the name of your repository:

```{bash}
echo \# fundamentals > README.md
cat README.md
```

### Git Status

Now that we actually have a file in our repository, we are ready to use Git. The one command that you'll find yourself using regularly in order to check the status of your project and see what changes have occurred since the last "clean" state is `git status`:

```{bash}
git status
```

Notice the response from our command. It lists our README.md file as an "untracked file". This is Git's way of telling you that a new file is present in the repository. It also says to use `"git add"` to track.

### Staging

Git allows you to review your changes before they get recorded into version control. This is called **staging**. You can add or remove files from the current staging area ("snapshot") using `git add` and `git rm`:

```{bash}
git add README.md
git status
```

Now, README.md is being tracked and is ready to be *committed* into version control.

### Git Commit

When you are ready to officially record ("commit") your staged changes, use the `git commit` command. Doing so will prompt you for a commit message (all commits get accompanied by a message), however, you can avoid the prompt and supply the message while requesting the commit by using the `-m` option:

```{bash}
git commit -m "initial commit"
git status
```

This will commit any staged files. Note that each commit is given a unique identifer known as a **SHA-1 hash**.

### Git Log

Use the `git log` command to print a summary of all the commits that you've made. This will include the full commit hash, author, date and message. For a more abbreviated result, use the `--oneline` option which will only print out the commit message and the first 7 characters from the commit hash:

```{bash}
git log
```

```{bash}
git log --oneline
```

### .gitignore

There may be instances when you want Git to ignore certain files in your repository. A good example of this is temporary data files that your code produces while it is executing. Generally, if you subscribe to the principles of reproducible research, you should be able to ignore any ouput files that your code produces as the code should be able to reproduce that output when necessary. For Git to ignore certain files, you'll need to create a `.gitignore` file in your project root, and in it list the file names or patterns that describe what is to be ignored.

Let's confirm a clean working directory:

```{bash}
ls -l
```
```{bash}
git status
```

Now, let's create a few data files that we'll want to ignore:

```{bash}
for i in 1 2 3; do
    touch data${i}.csv
done
ls -l
```
```{bash}
git status
```

Next, we simply create our `.gitignore` file with the correct pattern to ignore these new data files:

```{bash}
echo data\*.csv > .gitignore
cat .gitignore
```
```{bash}
git status
```

```{bash}
ls -la
```

Great! Git will now ignore any file that fits the pattern `data*.csv`. However, it does recognize that we've now introduced a new file - namely, the `.gitignore` file. So we simply add `.gitignore` to the staging area and commit:

```{bash}
git add .gitignore
git commit -m "added .gitignore"
git status
```
```{bash}
git log --oneline
```

### Discard Changes

Let's add a new file to our project:

```{bash}
echo "Hello, world!" > file1
cat file1
```

```{bash}
git add file1
git commit -m "added file1"
```
```{bash}
git log --oneline
```
```{bash}
git status
```

Now, this is where things start to heat up! Let's make a change to `file`:

```{bash}
echo "Goodbye, world:(" > file1
cat file1
```
```{bash}
git status
```

Based on the above response from `git status`, Git is aware that `file1` has changed! But, recall **staging**! This change has not yet been staged. Sometimes, changes like this might occur by accident or they may simply no longer be desired. To roll back unstaged changes to a file, use the `git checkout --` command:

```{bash}
git checkout -- file1
cat file1
```
```{bash}
git status
```

**Voila!** Everything is back to the way it was before we changed `file1`.

### Git Revert

In some cases, it will be necessary to undo an entire commit. To do so, use the `git revert` command and supply it with the SHA-1 hash of the commit that you would like to revert. Suppose we didn't want a `.gitignore` file. We would want to revert our second commit:

```{bash}
git log --oneline
```
```{bash, echo=FALSE}
git log --oneline | head -n 2 | tail -n 1 | cut -d " " -f 1 > ~/revert
echo '<span class="kw">git</span>' revert $(cat ~/revert) > ~/revert
```

`r htmltools::pre(class = "sourceCode bash", htmltools::code(class = "sourceCode bash", htmltools::includeHTML("~/revert")))`

```{bash, echo=FALSE}
git revert $(git log --oneline | head -n 2 | tail -n 1 | cut -d " " -f 1)
rm ~/revert
```

```{bash}
git status
```
```{bash}
git log --oneline
```
```{bash}
ls -la
```

Recall that the point of Git is to track ***all*** of your changes! Notice that `git revert` did not simply roll back or remove the prior commit, but rather it created a new commit reflective of the state we desired. In fact, by removing our `.gitignore` file, Git is now aware of the data files that it was previously ignoring! We can now *revert our revert* to get back our `.gitignore` file:

```{bash, echo=FALSE}
git log --oneline | head -n 1 | cut -d " " -f 1 > ~/revert
echo '<span class="kw">git</span>' revert $(cat ~/revert) > ~/revert
```

`r htmltools::pre(class = "sourceCode bash", htmltools::code(class = "sourceCode bash", htmltools::includeHTML("~/revert")))`

```{bash, echo=FALSE}
git revert $(git log --oneline | head -n 1 | cut -d " " -f 1)
rm ~/revert
```

```{bash}
git status
```
```{bash}
git log --oneline
```
```{bash}
ls -la
```

**AMAZING!!!**

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir='/home/dnegrey')
```
```{bash, echo=FALSE}
rm -rf fundamentals
```

\  
\  
\  

## Notes

### System Info

* **Operating system**: ubuntu 16.04 LTS (64-bit)
* **Bash version**: GNU bash, version 4.3.46(1)-release (x86_64-pc-linux-gnu)
* **Git version**: 2.7.4
